---
title: "Regularized multiple regression: drug sensitivity x gene vulnerability"
author: Nicholas Wang
output: html_document
runtime: shiny_prerendered
---

### Motivation
Building upon [Gonçalves et al.](https://www.biorxiv.org/content/10.1101/2020.01.14.905729v1.full), the following analysis extends their linear mixed models approach from single drug-gene pairs to regularized multiple regression of single drugs with all genes. The goal is to identify which gene knockouts best recapitulate the effects of individual drugs to elucidate  mechanism of action, unknown targets/side-effects, and small-molecule candidates with therapeutic potential.

### Data
Dataset           | File
----------------- | ---------------
GDSC2 drug data   | GDSC2_fitted_dose_response_15Oct19.xlsx  
Broad CRISPR data | Achilles_gene_effect.csv  
Cell line info    | sample_info.csv  

### Methods
glmnet R package  
LASSO regression with min lambda from cvfit

### Preliminary Results
```{r setup, include=FALSE}
library(tidyverse);
library(readxl);
library(glmnet);
library(ggplot2);
```

```{r data, cache=TRUE, include=FALSE}
# Import GDSC2 drug data
df.drug <- read_excel('../data/GDSC2_fitted_dose_response_15Oct19.xlsx') %>% 
  select(DRUG_NAME, COSMIC_ID, LN_IC50) %>% 
  distinct(DRUG_NAME, COSMIC_ID, .keep_all = TRUE) %>% 
  pivot_wider(names_from = DRUG_NAME, values_from = LN_IC50);

# Import Broad CRISPR data
df.gene <- read_csv('../data/Achilles_gene_effect.csv') %>% 
  drop_na();

# Cell line info
df.sample <- read_csv('../data/sample_info.csv', col_types = cols(additional_info = col_character())) %>% 
  select(DepMap_ID, stripped_cell_line_name, COSMIC_ID);
```

```{r echo=FALSE}
selectInput(
  inputId = 'drug',
  label = 'Drug',
  choices = colnames(df.drug)[-1]
  );

plotOutput('cvfit')
tableOutput('fit')
tableOutput('betas')
plotOutput('betas.plot')
```

```{r context = 'server', echo=FALSE}
lambdas <- 10^seq(2, -2, by = -0.1);

df.drug.1 <- reactive({
  merge(df.sample, select(df.drug, COSMIC_ID, input$drug), by = 'COSMIC_ID', all.y = TRUE) %>%
    merge(df.gene, by.x = 'DepMap_ID', by.y = 'X1') %>%
    filter(!is.na(input$drug)) %>%
    drop_na() %>%
    as_tibble()
  });

cvfit <- reactive({
  cv.glmnet(
    as.matrix(df.drug.1()[-(1:4)]),
    pull(df.drug.1(), input$drug),
    alpha = 1,
    lambda = lambdas
    );
  });

output$cvfit <- renderPlot({
  plot(cvfit());
  });

fit.min <- reactive({
  glmnet(
    as.matrix(df.drug.1()[-(1:4)]),
    pull(df.drug.1(), input$drug),
    alpha = 1,
    lambda = cvfit()$lambda.min
    );
  });

output$fit <- renderTable({
  print(fit.min())
})

betas.cor <- reactive({
  betas <- coef(fit.min()) %>%
    as.matrix() %>%
    as.data.frame() %>%
    rownames_to_column('gene') %>%
    filter(s0 > 0, gene != '(Intercept)') %>%
    arrange(desc(s0));
  cors <- cor(df.drug.1()[input$drug], df.drug.1()[betas$gene]) %>%
    t();
  out <- cbind(betas, cors) %>%
    rename(beta = s0, cor = input$drug);
  rownames(out) <- NULL;
  out
})

output$betas <- renderTable({
  betas.cor()
})

output$betas.plot <- renderPlot({
  ggplot(betas.cor(), aes(seq_len(length(beta)))) +
    geom_line(aes(y = beta, color = 'beta')) +
    geom_line(aes(y = cor, color = 'cor'))
})
```

### Future Directions
Adapting the model to fit single gene knockouts with the entire drug set to identify drugs or drug combinations best suited for targeting specific genes.



